---
title: "Bison_Project"
output: html_document
date: "2025-07-30"
---
load packages.tidyverse is a collection of R packages (like dplyr, ggplot2, readr, etc.) designed for data science.
```{r}
library(tidyverse)

```

load data
```{r}
veg_bison<- read.csv("../Data/knb-lter-knz/PVC021.csv")
veg_cattle <- read.csv("../Data/Cattle grazed watersheds/PBG011.csv")
Meteorological_data <- read.csv("../Data/Meteorological data from the konza prairie/AWE012.csv")
watersheds<- read.csv("../Data/Watersheds.csv")
```

```{r}
summary (veg_bison)
summary (veg_cattle)
```
```{r}
names(veg_bison)
names(veg_cattle)
```
```{r}
# Rename cattle-grazed dataset column names to match bison-grazed dataset
names(veg_cattle) <- c("DataCode", "RecType", "RecYear", "RecMonth", "RecDay",
                       "WaterShed", "SoilType", "Transect", "Plot", "SpeCode",
                       "AB_genus", "AB_species", "CoverClass", "Pid", "Comments")
```
```{r}
names(veg_bison)
names(veg_cattle)
```
```{r}
# Ensure both datasets have CoverClass
names(veg_bison)[13] <- "CoverClass"
names(veg_cattle)[13] <- "CoverClass"
#Combine both datasets into one
vegetation <- rbind(veg_bison, veg_cattle)
```
Take the watershed names in the column, and tell R to treat them as groups or labels, not as text.
```{r}
vegetation$WaterShed <- factor(vegetation$WaterShed)
```
This returns a vector of all the unique (non-duplicate) values in the WaterShed column of vegetation data frame.
```{r}
unique(vegetation$WaterShed)
```
```{r}
# Categorize watersheds
vegetation$grazing_treatment <- ifelse(str_starts(vegetation$WaterShed, "C"),"Cattle Grazed",
ifelse(str_starts(vegetation$WaterShed, "n"), "Bison Grazed", 
"Ungrazed"))
# Show unique watersheds with their treatment types
unique(vegetation[c("WaterShed", "grazing_treatment")])
```
Rename columns to match with Meteorological data
```{r}
# Rename columns to match
colnames(Meteorological_data)[colnames(Meteorological_data) == "RECYEAR"] <- "RecYear"
colnames(Meteorological_data)[colnames(Meteorological_data) == "RECMONTH"] <- "RecMonth"  
colnames(Meteorological_data)[colnames(Meteorological_data) == "RECDAY"] <- "RecDay"
```

```{r}
# Convert meteorological variables to numeric
Meteorological_data[c("TAVE", "DHUMID", "DSRAD", "DPPT")] <- lapply(Meteorological_data[c("TAVE", "DHUMID", "DSRAD", "DPPT")], as.numeric)

# Calculate yearly averages directly from meteorological data (months 4-10). create a yearly_temp dataframe with two colums; RecYear and Average of Meteorological data
#yearly_temp <- aggregate(cbind(TAVE,DHUMID,DSRAD,DPPT) ~ RecYear, 
#                        data = Meteorological_data[Meteorological_data$RecMonth >= 4 & Meteorological_data$RecMonth <= 10, ],
#                        FUN = mean, na.rm = TRUE)
yearly_temp2 <- Meteorological_data%>% 
  group_by(RecYear)%>%
  filter(RecMonth %in% seq(4,10,1)) %>% 
  summarise(mean_temp = mean(TAVE, na.rm= TRUE),
            mean_DHUMID = mean (DHUMID, na.rm = TRUE),
            mean_DSRAD = mean (DSRAD, na.rm = TRUE),
            mean_DPPT = mean (DPPT, na.rm = TRUE))
# Merge vegetation data with yearly averages only
merged_data <- merge(vegetation, yearly_temp2, by = "RecYear", all.x = TRUE)
```

```{r}
# Merge genus and species columns with space
merged_data$species_name <- paste(merged_data$AB_genus, merged_data$AB_species, sep = " ")
```

```{r}
# Add fire data to merged dataset by matching watershed names
merged_data <- merge(merged_data, 
                     watersheds[, c("WaterShed", "FireInterval_yr", "FireSeason","Description")],                           # select fire columns
                     by = "WaterShed",    # match on watershed name
                     all.x = TRUE)        # keep all vegetation records
```

```{r}
merged_data <- merged_data %>%
  mutate(
    scientific_name = paste(AB_genus, AB_species),
    pctCover = case_when(
      CoverClass == 1 ~ 0.5,
      CoverClass == 2 ~ 3.5,
      CoverClass == 3 ~ 15,
      CoverClass == 4 ~ 37.5,
      CoverClass == 5 ~ 62.5,
      CoverClass == 6 ~ 85,
      CoverClass == 7 ~ 97.5
    ),
    transect = case_match(
      Transect,
      "a" ~ "a", "A" ~ "a",
      "b" ~ "b", "B" ~ "b", 
      "c" ~ "c", "C" ~ "c",
      "d" ~ "d", "D" ~ "d"),
    plot_ID = paste0(WaterShed, Plot, transect, SoilType, RecYear, RecMonth, RecDay)
  )
```

```{r}
spp_matrix <- merged_data %>%
  filter(!(plot_ID == "004b1dt2024816" & 
           scientific_name == "teucri canad" & 
           CoverClass == 1)) %>%
  pivot_wider(id_cols = plot_ID,
              names_from = scientific_name,
              values_from = pctCover,
              values_fill = 0,
              values_fn = mean)
```


```{r}
merged_data$AB_genus<- NULL 
merged_data$AB_species<- NULL 
```

```{r}
plot_data <- merged_data %>%
  select(-scientific_name, -pctCover, -SpeCode, -CoverClass, -species_name) %>%
  distinct()

#final_data <- merge(diversityIndex, plot_data, by = "plot_ID", all.x = TRUE)
#Use inner_join to keep only matching records
final_data <- inner_join(diversityIndex, plot_data, by = "plot_ID")
```






